---
title: "How Busy Are Melbourne’s CBD Crossings? A Statistical Deep Dive"
description: "Modelling pedestrian flows at Southern Cross, Flinders Street, and QV Melbourne using probability models, bootstrapping, and confidence intervals."
author: "Jyovika Aswale"
date: 2 December 2025
categories: [R, Statistics, Pedestrian counts, Melbourne]
format:
  html:
    toc: true
    toc-location: left
    code-summary: "Show code"
    df-print: default
---

::: {.callout-note collapse="true" icon="people" title="Collaboration & Original Assignment"}
This analysis began as a group assignment for our Statistical Thinking unit, completed together with **Sia Chawla** and **Siddhi Jadhav**.  
The core data wrangling, modelling, and statistical code were developed collaboratively.

This blog post is my own adapted version for my analytics portfolio, I’ve rewritten the narrative, added explanations, and restructured the presentation for a broader readership, but the underlying work is **not solely my own**.

You can find the original group repository here:  
👉 **Original Assignment GitHub Repo:** <https://github.com/jyovika/Pedestrian_Crossings_Data_Analysis> 
:::

## Introduction 

The centre of Melbourne is full of movement and energy. Every day thousands of people pass through places like Southern Cross, Flinders Street, and QV Melbourne. At first glance it feels chaotic. Underneath that movement however is a pattern. I wanted to understand those patterns and tell a story about how these three locations behave.

I approached this like a real analytics problem. Start with descriptive exploration. Fit distributions that reflect the character of each location. Estimate meaningful quantities for planning and design. Compare sites in a way that answers the practical questions that planners and marketers actually care about.
By the end of the analysis, these three crossings no longer felt like random busy spots on a map. They felt like distinct personalities with their own rhythms and their own commercial and operational implications.

```{r, setup, echo=FALSE}
#| label: setup
#| include: false
#| message: false
#| warning: false
knitr::opts_chunk$set(
fig.align = "center",
fig.width = 8,
fig.height = 5,
dpi = 150,
echo = TRUE,
warning = FALSE,
message = FALSE
)

library(boot)
library(MASS)
library(tidyverse)
library(knitr)
library(kableExtra)
library(broom)
library(purrr)
library(scales)

set.seed(355435)

cross_cols <- c(
"Flinders Street" = "#766153",
"QV Melbourne" = "#718f94",
"Southern Cross"  = "#717744")

```

```{r,cleaning-and-loading-data, echo=FALSE}
pedestrian_df <- read.csv(here::here("data/pedestrian/pedestrians.csv"))

pedestrian_long <- pedestrian_df %>%
  mutate(obs = row_number()) %>%
  pivot_longer(-obs, names_to = "crossing", values_to = "count") %>%
  mutate(crossing = recode(crossing,
                           "flinders_street" = "Flinders Street",
                           "qv_melbourne"    = "QV Melbourne",
                           "southern_cross"  = "Southern Cross")) %>%
  mutate(crossing = factor(
    crossing,
    levels = c("Southern Cross", "Flinders Street", "QV Melbourne")))
```

## First Impressions Through Visuals

To get a feel for how each location behaves, I plotted boxplots and density curves. These gave a surprisingly clear first impression.

::: columns
::: {.column width="50%"}

```{r, echo=FALSE}
#| label: fig-boxplot
#| fig-cap: "Pedestrian counts by crossing."
ggplot(pedestrian_long, aes(x = crossing, y = count, fill = crossing)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  scale_fill_manual(values = cross_cols) +
  labs(x = NULL, y = "Count per interval") +
  theme_minimal() +
  theme(legend.position = "none")

```

:::
::: {.column width="50%"}

```{r, echo=FALSE}
#| label: fig-density
#| fig-cap: "Distribution of pedestrian counts by crossing."
ggplot(pedestrian_long, aes(x = count, fill = crossing)) +
  geom_density(alpha = 0.35) +
  facet_wrap(~ crossing, scales = "fixed") +
  scale_fill_manual(values = cross_cols) +
  labs(x = "Count", y = "Density") +
  theme_minimal() +
  theme(legend.position = "none")
```

:::
:::

QV Melbourne clearly carries the heaviest flow. Southern Cross and Flinders Street sit closer together although Flinders has slightly lower typical values. All three distributions show right skew which tells us they have occasional high traffic bursts.

The summary statistics confirm these observations.
```{r, echo=FALSE}
#| label: tbl-crossing-summary
pedestrian_long %>%
  group_by(crossing) %>%
  summarise(
    n      = n(),
    mean   = mean(count),
    median = median(count),
    sd     = sd(count),
    min    = min(count),
    max    = max(count),
    iqr    = IQR(count),
    .groups = "drop"
  ) %>%
  kable(
    digits = 1,
    col.names = c("Crossing", "n", "Mean", "Median", "SD", "Min", "Max", "IQR")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE
  ) %>%
  row_spec(0, bold = TRUE, color = "#FFFFFF", background = "#598392") %>%
  column_spec(1, bold = TRUE, color = "#004D40")
```

At this stage it already felt like QV Melbourne was in a different league. It is not just busy. It is volatile. That volatility becomes important once we start modelling.





